Semantic integration of physiology phenotypes: an application to the cellular phenotype ontology

### 2012-01-19 trunk/paper/drinking-bioinformatics.tex

\documentclass{bioinfo}
\usepackage{setspace}

\copyrightyear{2005}
\pubyear{2005}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{multirow}
\usepackage{pifont}
% Load packages
\usepackage{cite} % Make references as [1-4], not [1,2,3,4]
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage[utf8]{inputenc} %unicode support
\urlstyle{rm}
\usepackage{natbib}
\renewcommand{\cite}{\citep}
\usepackage{amsmath}
\usepackage{amssymb}

\usepackage{pifont}
%\newcommand{\tick}{\ding{52}}
\newcommand{\tick}{\checkmark}
\newcommand{\tickNo}{\hspace{1pt}\ding{55}}

@@

\title{Semantic integration of physiology phenotypes: an application
  to the cellular phenotype ontology}

\author{Robert Hoehndorf$^1$ and Midori
  A. Harris$^2$ and Heinrich Herre$^3$ and Gabriella Rustici$^4$ and
  Georgios V. Gkoutos$^{1}$}

@@

@@

\begin{abstract}
  The systematic observation of phenotypes has become a crucial tool
  of functional genomics, and several large international projects are
  currently underway to identify and characterize the phenotypes that
  are associated with genotypes in several species. To integrate
  phenotype descriptions within and across species, phenotype
  ontologies have been developed. Applying ontologies to unify
  phenotype descriptions in the domain of physiology has been a
  particular challenge due to the high complexity of the underlying
  domain.  Here, we present the outline of a theory for an ontology of
  physiology-related phenotypes. We provide a formal description of
  process attributes and relate them to the attributes of their parts
  and participants. We apply our theory to create the Cellular
  Phenotype Ontology (CPO). The CPO is an ontology of morphological
  and physiological phenotypic abnormalities of cells, cell components
  and cellular processes. Its prime application is the unification of
  cellular phenotype descriptions across species by providing terms
  and uniform definition patterns. The CPO can further be used for the
  annotation of observed abnormalities in domains, such as systems
  microscopy, in which cellular abnormalities are observed and for
  which no phenotype ontology has been created.  The CPO and the
  source code we generated to create the CPO are available on
  \url{http://cell-phenotype.googlecode.com}.
\end{abstract}

{\bf Keywords:} cell phenotype, process, ontology, Semantic Web

@@

\section{Materials and Methods}
\subsection{Formal ontology}
The ontology as an approach to semantic standardisation was proposed
more than a decade ago and since then has become the dominant
methodology used to semantically categorise phenodeviance.  The
biomedical research community has invested considerable effort and
resources in the development and establishment of ontologies that are
becoming increasingly successful as information management and
integration tools in many disparate scientific fields allowing
interoperability and semantic information processing between diverse
biomedical resources and domains.

@@

\subsection{Processes and their participants}
\begin{figure}
  \centering
  \includegraphics[width=.45\textwidth]{onto.pdf}
  \caption{Overview over basic top-level categories. Physical objects
    are entities that are wholly present at time points. Qualities are
    attributes of entities. Functions are capabilities that arise from
    physical objects with their qualities, and processes are
    temporally extended entities that may be realizations of a
    function.\label{fig:onto}}
\end{figure}

Most biomedical ontologies share common distinctions between different
kinds of entities: physical objects, qualities, functions and
processes. A physical object is an entity that is present as a whole
at a time point, i.e., an entity that has no temporal parts. A quality
is an attribute or feature of an entity. Physical objects together
with their qualities give rise to functions, which are capabilities or
potentials of physical objects. These functions can then be realized
by processes, which are temporally extended entities
\cite{Burek2006}. Examples of classes that may have processes as
instances include {\em Drinking}, {\em Triathlon} or {\em Apoptosis}.
Figure \ref{fig:onto} illustrates these basic categories of being.

@@

@@

\begin{figure}
  \centering
  \includegraphics[width=.5\textwidth]{processpatterns.pdf}

@@

\subsection{Cell Phenotype Ontology}
\begin{figure*}
  \centering
  \includegraphics[width=\textwidth,
  height=.33\textheight]{overview.pdf}
  \caption{Overview over the taxonomic structure of CPO. The structure
    is based on a cellular component class $X$ and the cellular
    processes $P(X)$ in which $X$ is involved.\label{fig:overview}}
\end{figure*}

While our considerations about process attributes are only the
beginnings of a full-fledged theory, we have derived several phenotype
formalization patterns and a high-level taxonomic structure of
process-based phenotypes. To evaluate our approach, we created the
Cellular Phenotype Ontology (CPO) by automatically applying our
patterns to the GO.

@@

\begin{table*}
  \centering
  \begin{tabular}{l|l|l|l}
    & Increased cytoplasmic flow rate & Normal cytoplasmic flow rate & Decreased
    cytoplasmic flow rate \\
    \hline
    Increased frequency &increased total flow rate &increased total
    flow rate &?\\
    Normal frequency &increased total flow rate &normal total flow
    rate &decreased total flow rate\\
    Decreased frequency &?&decreased total flow rate &decreased total
    flow rate\\
    \hline
  \end{tabular}
  \caption{\label{tbl:flow}Interdependency for the attribute {\em
      Total cytoplasmic flow rate}. A {\em Total cytoplasmic flow
      rate} is an attribute of {\em Regulation of cytoplasmic
      streaming} processes, while {\em Cytoplasmic flow rate} is an
    attribute of individual {\em cytoplasmic streaming}
    processes. Depending both on whether the cytoplasmic flow rate in
    individual {\em cytoplasmic streaming} processes is increased or
    decreased and whether the frequency of occurrence of {\em
      cytoplasmic streaming} is increased or decreased, the total
    cytoplasmic flow rate can be increased or decreased.}
\end{table*}

@@

\begin{table*}
  \centering
  \begin{tabular}{|p{7cm}|l|p{4cm}|}
    Neumann et al.\cite{Neumann2010}&Schmitz et
    al.\cite{Schmitz2010}&Fuchs et al.\cite{Fuchs2010}\\
    \hline
    binuclear & normal mitotic exit & actin fiver cells\\
    cell death & prolonged mitotic exit & big cells \\
    cell migration & & bright and large cells phenotype \\
    condensation followed by decondensation without completion of
    mitosis & & bright nuclei\\
    condensation without mitosis/collapse of nucleus & & cells with protrusions\\
    dynamic changes & & elongated cells\\
    failure in decondensation & & elongated cells with protrusions\\
    grape& & high actin ratio cells \\
    increased proliferation& & lamellipodia + high actin ratio cells\\
    large& & lamellipodia cells\\
    large nucleus& & large cells\\
    metaphase alignment problems/including no metaphase& & large nuclei\\
    metaphase delay/arrest& & low eccentricity cells\\
    migration (distance)& & metaphase cells\\
    migration (speed)& & proliferating cells\\
    mitotic delay/arrest& & small cells\\
    nuclei stay close together& & small cells with an enrichment of mitotic cells\\
    polylobed& &\\
    pulsating nuclei& &\\
    segregation problems/chromatin bridges/lagging chromosomes/multiple DNA masses&&\\
    small nucleus&&\\
    strange nuclear shape&&\\
    \hline
  \end{tabular}
  \caption{\label{tbl:studies}The table summarizes cellular phenotype
    terms used in three recent systems microscopy studies.}
\end{table*}

@@

\section{Discussion}

\subsection{Applications of the CPO}
The Fission Yeast Phenotype Ontology (FYPO), a new ontology developed
to support annotation of phenotypes in {\em Schizosaccharomyces
  pombe}, consists of pre-composed terms describing normal or abnormal
cellular phenotypes. Over 80\% of FYPO definitions reference
descendants of GO-BP's {\em Cellular process} as the entity; a further
11\% reference GO-CC terms. All FYPO explicit definitions reference
qualities in PATO, including {\em normal}, {\em abnormal}, and several
process qualities including {\em increased duration} and {\em
  decreased occurrence}. FYPO will thus fit neatly under the CPO
umbrella, and stands to benefit from the automated synchronization
between CPO and GO, as well as the integration of cellular phenotypes
across species that the CPO can provide. {\em Schizosaccharomyces
  pombe} annotations to FYPO terms will provide a rich body of highly
specific, well-supported data to be integrated with data from other
species.

@@

\section{Acknowledgements}
Funding for RH is provided by the European Commission's 7th Framework
Programme, RICORDO project, grant number 248502. MAH is supported by
Wellcome Trust, grant WT090548MA. HH is supported by the Institute for
Medical Informatics, Statistics and Epidemiology, University of
Leipzig.  Funding for GR is provided by the European Union's Seventh
Framework Programme (FP7/2007-2013) under grant agreement number
258068, EU-FP7-Systems Microscopy NoE. 
%GVG is funded by NIH grant number R01 HG004838-02.

% % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \section*{Authors contributions}

% % %%%%%%%%%%%%%%%%%%%%%%%%%%%
% The authors declare that they have no competing interests.

% % %%%%%%%%%%%%%%%%%%%%%%%%%%%
% \section*{Acknowledgements}
% \ifthenelse{\boolean{publ}}{\small}{}

% The research work in its first unrevised form was presented at the
% SMBM 2010, Hinxton, Cambridge, U.K.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The Bibliography %%
%%                                                         %%
%% Bmc_article.bst will be used to %%
%% create a .BBL file for submission, which includes %%
%% XML structured for BMC.  %%
%%                                                         %%
%%                                                         %%
%% Note that the displayed Bibliography will not %%
%% necessarily be rendered by Latex exactly as specified %%
%% in the online Instructions for Authors.  %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliographystyle{plain} % Style BST file
\bibliography{lc}

>>>>

\documentclass{bioinfo}
\copyrightyear{2005}
\pubyear{2005}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{multirow}
\usepackage{pifont}
%\newcommand{\tick}{\ding{52}}
\newcommand{\tick}{\checkmark}
\newcommand{\tickNo}{\hspace{1pt}\ding{55}}

@@

\title[Cell Phenotype Ontology]{Semantic integration of physiology
  phenotypes: an application to the cellular phenotype ontology}

\author[Hoehndorf \textit{et~al}]{Robert Hoehndorf$^{1}$\footnote{to
    whom correspondence should be addressed}, Midori
  A. Harris$^2$, Heinrich Herre$^3$, Gabriella Rustici$^4$ and
  Georgios V. Gkoutos$^{1}$}

@@

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

@@

\begin{abstract}
\section{Motivation:}
The systematic observation of phenotypes has become a crucial tool of
functional genomics, and several large international projects are
currently underway to identify and characterize the phenotypes that
are associated with genotypes in several species. To integrate
phenotype descriptions within and across species, phenotype ontologies
have been developed. Applying ontologies to unify phenotype
descriptions in the domain of physiology has been a particular
challenge due to the high complexity of the underlying domain.

\section{Results:}
Here, we present the outline of a theory for an ontology of
physiology-related phenotypes. We provide a formal description of
process attributes and relate them to the attributes of their parts
and participants. We apply our theory to create the Cellular Phenotype
Ontology (CPO). The CPO is an ontology of morphological and
physiological phenotypic abnormalities of cells, cell components and
cellular processes. Its prime application is the unification of
cellular phenotype descriptions across species by providing terms and
uniform definition patterns. The CPO can further be used for the
annotation of observed abnormalities in domains, such as systems
microscopy, in which cellular abnormalities are observed and for which
no phenotype ontology has been created.

\section{Availability and implementation:} 
The CPO and the source code we generated to create the CPO are freely
available on \url{http://cell-phenotype.googlecode.com}.

\section{Contact:} \href{rh497@cam.ac.uk}{rh497@cam.ac.uk}
\end{abstract}

@@

\begin{methods}
\section{System and methods}
\subsection{Formal ontology}
The ontology as an approach to semantic standardisation was proposed
more than a decade ago and since then has become the dominant
methodology used to semantically categorise phenodeviance.  The
biomedical research community has invested considerable effort and
resources in the development and establishment of ontologies that are
becoming increasingly successful as information management and
integration tools in many disparate scientific fields allowing
interoperability and semantic information processing between diverse
biomedical resources and domains.

@@

\subsection{Processes and their participants}
Most biomedical ontologies share common distinctions between different
kinds of entities: physical objects, qualities, functions and
processes. A physical object is an entity that is present as a whole
at a time point, i.e., an entity that has no temporal parts. A quality
is an attribute or feature of an entity. Physical objects together
with their qualities give rise to functions, which are capabilities or
potentials of physical objects. These functions can then be realized
by processes, which are temporally extended entities
\cite{Burek2006}. Examples of classes that may have processes as
instances include {\em Drinking}, {\em Triathlon} or {\em Apoptosis}.
Figure \ref{fig:onto} illustrates these basic categories of being.

@@

\end{methods}

@@

@@

\subsection{Cell Phenotype Ontology}
While our considerations about process attributes are only the
beginnings of a full-fledged theory, we have derived several phenotype
formalization patterns and a high-level taxonomic structure of
process-based phenotypes. To evaluate our approach, we created the
Cellular Phenotype Ontology (CPO) by automatically applying our
patterns to the GO.

@@

\begin{table*}
  \centering
  \begin{tabular}{p{2.8cm}|p{4cm}|p{4cm}|p{4cm}}
    & Increased flow rate & Normal flow rate & Decreased
    flow rate \\
    \hline
    Increased frequency &increased total flow rate &increased total
    flow rate &?\\
    Normal frequency &increased total flow rate &normal total flow
    rate &decreased total flow rate\\
    Decreased frequency &?&decreased total flow rate &decreased total
    flow rate\\
    \hline
  \end{tabular}
  \caption{\label{tbl:flow}Interdependency for the attribute {\em
      Total cytoplasmic flow rate}. A {\em Total cytoplasmic flow
      rate} is an attribute of {\em Regulation of cytoplasmic
      streaming} processes, while {\em Cytoplasmic flow rate} is an
    attribute of individual {\em cytoplasmic streaming}
    processes. Depending both on whether the cytoplasmic flow rate in
    individual {\em cytoplasmic streaming} processes is increased or
    decreased and whether the frequency of occurrence of {\em
      cytoplasmic streaming} is increased or decreased, the total
    cytoplasmic flow rate can be increased or decreased.}
\end{table*}

@@

% \begin{table*}
%   \centering
%   \begin{tabular}{|p{7cm}|l|p{4cm}|}
%     Neumann et al.\cite{Neumann2010}&Schmitz et
%     al.\cite{Schmitz2010}&Fuchs et al.\cite{Fuchs2010}\\
%     \hline
%     binuclear & normal mitotic exit & actin fiver cells\\
%     cell death & prolonged mitotic exit & big cells \\
%     cell migration & & bright and large cells phenotype \\
%     condensation followed by decondensation without completion of
%     mitosis & & bright nuclei\\
%     condensation without mitosis/collapse of nucleus & & cells with protrusions\\
%     dynamic changes & & elongated cells\\
%     failure in decondensation & & elongated cells with protrusions\\
%     grape& & high actin ratio cells \\
%     increased proliferation& & lamellipodia + high actin ratio cells\\
%     large& & lamellipodia cells\\
%     large nucleus& & large cells\\
%     metaphase alignment problems/including no metaphase& & large nuclei\\
%     metaphase delay/arrest& & low eccentricity cells\\
%     migration (distance)& & metaphase cells\\
%     migration (speed)& & proliferating cells\\
%     mitotic delay/arrest& & small cells\\
%     nuclei stay close together& & small cells with an enrichment of mitotic cells\\
%     polylobed& &\\
%     pulsating nuclei& &\\
%     segregation problems/chromatin bridges/lagging chromosomes/multiple DNA masses&&\\
%     small nucleus&&\\
%     strange nuclear shape&&\\
%     \hline
%   \end{tabular}
%   \caption{\label{tbl:studies}The table summarizes cellular phenotype
%     terms used in three recent systems microscopy studies.}
% \end{table*}

@@

\section{Discussion}
\subsection{Applications of the CPO}
The Fission Yeast Phenotype Ontology (FYPO), a new ontology developed
to support annotation of phenotypes in {\em Schizosaccharomyces
  pombe}, consists of pre-composed terms describing normal or abnormal
cellular phenotypes. Over 80\% of FYPO definitions reference
descendants of GO-BP's {\em Cellular process} as the entity; a further
11\% reference GO-CC terms. All FYPO explicit definitions reference
qualities in PATO, including {\em normal}, {\em abnormal}, and several
process qualities including {\em increased duration} and {\em
  decreased occurrence}. FYPO will thus fit neatly under the CPO
umbrella, and stands to benefit from the automated synchronization
between CPO and GO, as well as the integration of cellular phenotypes
across species that the CPO can provide. {\em Schizosaccharomyces
  pombe} annotations to FYPO terms will provide a rich body of highly
specific, well-supported data to be integrated with data from other
species.

@@

\begin{figure}
  \centering
  \includegraphics[width=1\textwidth]{onto.pdf}
  \caption{Overview over basic top-level categories. Physical objects
    are entities that are wholly present at time points. Qualities are
    attributes of entities. Functions are capabilities that arise from
    physical objects with their qualities, and processes are
    temporally extended entities that may be realizations of a
    function.\label{fig:onto}}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=1\textwidth]{processpatterns.pdf}

\begin{figure*}
  \centering
  \includegraphics[width=1.2\textwidth, height=.4\textheight, angle=270]{overview.pdf}
  \caption{Overview over the taxonomic structure of CPO. The structure
    is based on a cellular component class $X$ and the cellular
    processes $P(X)$ in which $X$ is involved.\label{fig:overview}}
\end{figure*}

\section{Acknowledgement}
Funding for RH is provided by the European Commission's 7th Framework
Programme, RICORDO project, grant number 248502. MAH is supported by
Wellcome Trust, grant WT090548MA. HH is supported by the Institute for
Medical Informatics, Statistics and Epidemiology, University of
Leipzig.  Funding for GR is provided by the European Union's Seventh
Framework Programme (FP7/2007-2013) under grant agreement number
258068, EU-FP7-Systems Microscopy NoE.

\bibliographystyle{plain}
\bibliography{lc}
